COVID19 - Forecast analysis - time dependent SEIR model - Veneto

### 2020-03-14 draft_analysis_Italy.Rmd
Add home page, navigation and theme

---
title: "COVID19 - Forecast analysis"
author: "PG"
date: "3/13/2020"
output: github_document
---

@@

#estimate r0
#see https://kingaa.github.io/clim-dis/parest/parest.html
#calculate r0 based in the last XX observation 
```{r R0 trend}
beta_vec<-NULL
sd_vec<-NULL
for (i in 1:(days-6)){
fit <- lm(log(totale_attualmente_positivi)~t,data=dat_csv[i:days,])
beta_vec<-c(beta_vec,coef(fit)[2])
sd_vec<-c(sd_vec,coef(summary(fit))[2,2])
}

label <-days:7
mean  <- (beta_vec*14+1)
lower <- ((beta_vec-1.96*sd_vec)*14+1)
upper <- ((beta_vec+1.96*sd_vec)*14+1)

@@

library(ggplot2)
fp <- ggplot(data=df, aes(x=label, y=mean, ymin=lower, ymax=upper)) +
  geom_pointrange() + 
  geom_hline(yintercept=1, lty=2) +  # add a dotted line at x=1 after flip
  coord_flip() +  # flip coordinates (puts labels on y axis)
  xlab("Last days used for the calculation") + ylab("R0 Mean (95% CI)") +
  theme_bw()  # use a white background
print(fp)
```

- <img src="https://latex.codecogs.com/gif.latex?log(Y_t) = a + b*t + e_t" />

where $`Y_t`$ is the cumulative number of infected at the time t, while b is beta, the slope of the regression line.  
We calculate several R0 values, each one based on a different number of days before the last day. 
The R0 shows a decreasing trend in the last period. The slope b indicates the rate of exponetial increase.  
However we considere the value of the last 10 days, because is more stable.

R code:  
```{r data}
head(dat_csv)
fit1 <- lm(log(totale_attualmente_positivi)~t,data=dat_csv[(days-10):days,])
summary(fit1)
```

The slope coefficient estimated in the linear regression model can be used to estimate R0.

R0=1+b*incubation period.

The incubation period for the coronavirus is in mean 5.1 days with a range from 2-14 days.
Please see https://www.worldometers.info/coronavirus/coronavirus-incubation-period/.
In the calculation we considered an incubation period of 14 days for two reasons: 1) the majority of cases is asymptomatic, contagiousness is greater than 5, maybe 14. A minority (who made the swab) will have a duration of about 5 days between the start of contagiousness and swab; 2) 14 days is the worst scenario to consider in this phase.
```{r r0 plot, echo=TRUE}
slope <-coef(summary(fit1))[2,1]; slope
slope.se <- coef(summary(fit1))[2,2]; slope.se
### R0 estimates and 95% IC 
R_0=slope*14+1;R_0
(slope+c(-1,1)*1.96*slope.se)*14+1
```
We want to make a short term forecast (14 days) with 3 scenario:

-Scenario 1: 10 exposed people for each COVID-19 case and beta the same (no restrictions made or even no effects)

-Scenario 2: 10 exposed people for each COVID-19 case and beta reduced of 50% (-50% exposed people)

-Scenario 3: 5 exposed people for each COVID-19 case and beta reduced of 50%  (-50% both exposed people and COVID19 contagious power)

We fix a series of initial parameters:
- I0: initial number of COVID-19 cases  
- R0: initial number of recovered  
- beta: the quantity connected to R0  
- N: Italian population   
- duration: infection duration of COVID-19   
- sigma0: the coronavirus transmission rate (half of flu epidemic)  
- mu0: the overall mortality rate

```{r initation plot, echo=TRUE}
# initial number of infectus
I0<-dat_csv$totale_attualmente_positivi[dim(dat_csv)[1]]; I0
# initial number of recovered
R0<-dat_csv$dimessi_guariti[dim(dat_csv)[1]]; R0
#beta 
beta0<-R_0/(14)
# italian poulation
N=60480000
# duration of COVID19 
duration<-14
#sigma0 is the coronavirus transmission rate fixed to 5%  (half of flu epidemic)
sigma0<-0.05
#mortality rate 
mu0<-1/(82*365.25) # 1/lifespan
```
We use the library(EpiDynamics) and the function SEIR() to implement a SEIR model:

<img src="http://www.public.asu.edu/~hnesse/classes/seireqn.png"/>

```{r  scenario plot, echo=TRUE}
library(EpiDynamics)

# average number of single connections of an infected person
# less contacts, less probability of new infections
# we keep constant the other parameters
forecast<-14
parameters <- c(mu = mu0, beta = beta0, sigma = sigma0, gamma = 1/duration)
f1<-10
initials <- c(S = 0.95, E = (f1*I0/N), I = I0/N, R = R0/N)
seir1 <- SEIR(pars = parameters, init = initials, time = 0:forecast)
parameters <- c(mu = mu0, beta = beta0, sigma = sigma0, gamma = 1/duration)
f2<-5
initials <- c(S = 0.95, E = (f2*I0/N), I = I0/N, R = R0/N)
seir2 <- SEIR(pars = parameters, init = initials, time = 0:forecast)
parameters <- c(mu = mu0, beta = beta0*1/2, sigma = sigma0, gamma = 1/duration)
f3<-5
initials <- c(S = 0.95, E = (f3*I0/N), I = I0/N, R = R0/N)
seir3 <- SEIR(pars = parameters, init = initials, time = 0:forecast)

date<-seq(as.Date("2020-02-24"),as.Date("2020-02-24")+forecast-1+dim(dat_csv)[1],1)
plot(date,c(dat_csv$totale_attualmente_positivi,seir1$results$I[-1]*N),type="l",ylab="Cases",xlab="time",main="Infected")
lines(date,c(dat_csv$totale_attualmente_positivi,seir2$results$I[-1]*N),col=2)
lines(date,c(dat_csv$totale_attualmente_positivi,seir3$results$I[-1]*N),col=3)
lines(date[1:dim(dat_csv)[1]],dat_csv$totale_attualmente_positivi,lwd=2)
legend("topleft",c("first scenario","second scenario","third scenario"),lty=1,col=1:3)
```

The 3 scenarios show different numbers. 
If we consider the second scenario, at the end of the 2 weeks (`r date[length(date)]`) the number of infected is  (`r (seir2$results$I*N)[length(seir2$results$I)]`).

In the next plot the cumulative number of infected.  
At the end of the 2 weeks (`r date[length(date)]`) the total number of COVID19 cases is expected to be  (`r ((seir2$results$I+seir3$results$R)*N)[length(seir2$results$I)]`).

@@

plot(date,c(dat_csv$totale_casi,(seir1$results$I[-1]+seir1$results$R[-1])*N),type="l",ylab="Cases",xlab="time",main="Cumulative Infected")
lines(date,c(dat_csv$totale_casi,(seir2$results$I[-1]+seir2$results$R[-1])*N),col=2)
lines(date,c(dat_csv$totale_casi,(seir3$results$I[-1]+seir3$results$R[-1])*N),col=3)
lines(date[1:dim(dat_csv)[1]],(dat_csv$totale_casi),lwd=2)
legend("topleft",c("first scenario","second scenario","third scenario"),lty=1,col=1:3)
```

We make an effort to estimate the effect of restriction measures took by the Italian Government (11 March 2020). 
After 14 days  (`r date[length(date)]`) we have the current status in the second scenario:  
- I14=(`r (seir2$results$I*N)[length(seir2$results$I)]`);  
- R14=(`r (seir2$results$R*N)[length(seir2$results$R)]`);  
- S14=(`r (seir2$results$S*N)[length(seir2$results$S)]`);  
- E14=(`r (seir2$results$E*N)[length(seir2$results$E)]`);

We reduce R0 to 1.10 and the number of exposed to the number of infected.
We estimate the trend of the number infected, recovered and the total cases for other 14 days forward.

```{r  scenario plot2, echo=TRUE}
I14=seir2$results$I[length(seir2$results$I)]
R14=seir2$results$R[length(seir2$results$I)]
S14=seir2$results$S[length(seir2$results$I)]
E14=seir2$results$E[length(seir2$results$I)]
forecast2<-14
beta14<-1.1/14
parameters <- c(mu = mu0, beta = beta14, sigma = sigma0, gamma = 1/duration)
initials <- c(S = 1-I14*2-R14, E = I14, I = I14, R = R14)
seir2_2 <- SEIR(pars = parameters, init = initials, time = 0:forecast2)

date<-seq(as.Date("2020-02-24"),as.Date("2020-02-24")+forecast2+forecast-1+dim(dat_csv)[1],1)
plot(date,c(dat_csv$totale_attualmente_positivi,seir2$results$I[-1]*N,seir2_2$results$I[-1]*N),type="l",ylab="Cases",xlab="time",ylim=c(0,((seir2_2$results$I+seir2_2$results$R[-1])*N)[forecast2]),col=2)
lines(date,c(dat_csv$dimessi_guariti,seir2$results$R[-1]*N,seir2_2$results$R[-1]*N),col=3)
lines(date,c(dat_csv$totale_casi,(seir2$results$I[-1]+seir2$results$R[-1])*N,c(seir2_2$results$I[-1]+seir2_2$results$R[-1])*N),col=1)

legend("topleft",c("All cases","Infected","Recovered"),lty=1,col=1:3)
```

>>>>

---
title: "COVID19 - Forecast and predictions using a time dependent SEIR model" 
author: "PG"
date: "3/13/2020"
output: github_document
---

@@

#estimate r0
#see https://kingaa.github.io/clim-dis/parest/parest.html
#calculate r0 based in the last XX observation 
```{r R0 trend}
beta_vec<-NULL
sd_vec<-NULL
for (i in 1:(days-4)){
fit <- lm(log(totale_attualmente_positivi)~t,data=dat_csv[i:days,])
beta_vec<-c(beta_vec,coef(fit)[2])
sd_vec<-c(sd_vec,coef(summary(fit))[2,2])
}

label <-days:5
mean  <- (beta_vec*14)
lower <- ((beta_vec-1.96*sd_vec)*14)
upper <- ((beta_vec+1.96*sd_vec)*14)

@@

library(ggplot2)
fp <- ggplot(data=df, aes(x=label, y=mean, ymin=lower, ymax=upper)) +
  geom_pointrange() + 
  geom_hline(yintercept=1, lty=2) +  # add a dotted line at x=1 after flip
  coord_flip() +  # flip coordinates (puts labels on y axis)
  xlab("Last days used for the calculation") + ylab("R0 Mean (95% CI)") +
  theme_bw()  # use a white background
print(fp)

Y_t= a + beta * t +e_t

where $`Y_t`$ is the cumulative number of infected at the time t, while b is beta, the slope of the regression line.  
The slope coefficient estimated in the linear regression model can be used to estimate R0.

R0=slope*incubation period.

The incubation period for the coronavirus is in mean 5.1 days with a range from 2-14 days.
Please see https://www.worldometers.info/coronavirus/coronavirus-incubation-period/, but the incubation period is used for epidemic diseases that causes the immediate isolation of the infected subject.

In the calculation we considered an "incubation period" of 14 days for two reasons: 1) the majority of cases is asymptomatic, contagiousness is greater than 5, maybe 14. A minority (who made the swab) will have a duration of about 5 days between the start of contagiousness and swab; 2) 14 days is the worst scenario to consider in this phase.

We calculate several R0 values, each one based on a different number of days before the last day. 
The R0 shows a decreasing trend in the last period. The slope b indicates the rate of exponetial increase.  
However,R0 is going to decrease in the next days. We use the estimated trend between R0 and time to calculate the future R0 value for the next 14 days. 
We predict beta (and R0) for the next 14 days assuming a Gamma distribution for the beta (the slope) forcing is a value greater than 0. 
```{r R0 forecast}
library(splines)
time<-1:(days-4)

beta.model<-glm(beta_vec~bs(time,3),weights = 1/sd_vec,family=Gamma)
forecast=14
# add 'fit', 'lwr', and 'upr' columns to dataframe (generated by predict)
pre<-predict(beta.model,type='response',newdata=data.frame(time=1:(days+forecast)),se.fit=TRUE)
date<-seq(as.Date("2020-02-24"),as.Date("2020-02-24")+forecast-1+dim(dat_csv)[1],1)
beta.predict <- data.frame(beta_vec=c(beta_vec,rep(NA,forecast+4)),time=date,fit=pre$fit,lwr=pre$fit-1*1.96*pre$se.fit,upr=pre$fit+1*1.96*pre$se.fit)

r0.predict<-beta.predict
r0.predict[,c(1,3:5)]<-r0.predict[,c(1,3:5)]*14
# plot the points (actual observations), regression line, and confidence interval
p <- ggplot(r0.predict, aes(date,beta_vec))
p <- p + geom_point() +labs(x="time from t0",y="R0 value") 
p <- p + geom_line(aes(date,fit))
p <- p + geom_ribbon(aes(ymin=lwr,ymax=upr), alpha=0.3)
p
```

R0 passes from values greater than 4 in the initial phase to values <1 at the ending of the 14-days forecast.

We want to make a short term forecast (14 days) with 3 scenario:

-Scenario 1: 10 exposed people for each COVID-19 case (no restrictions made or even no effects)

-Scenario 2: 5 exposed people for each COVID-19 case (-50% exposed people)

-Scenario 3: 3 exposed people for each COVID-19 case  (-70% exposed people)

We use a SEIR model fixing a series of initial parameters:

- I0: initial number of COVID-19 cases  
- R0: initial number of recovered  
- beta: the quantity connected to R0 is considered to vary according the previous estimation
- N: Italian population   
- duration: infection duration of COVID-19   
- sigma0: the coronavirus transmission rate (half of flu epidemic)  
- mu0: the overall mortality rate

```{r initation plot, echo=TRUE}
# initial number of infectus
I0<-dat_csv$totale_attualmente_positivi[dim(dat_csv)[1]]; I0
# initial number of recovered
R0<-dat_csv$dimessi_guariti[dim(dat_csv)[1]]; R0
# italian poulation
N=60480000
# duration of COVID19 
duration<-14
#sigma0 is the coronavirus transmission rate fixed to 5%  (half of flu epidemic)
sigma0<-0.05
#mortality rate 
mu0<-1/(82*365.25) # 1/lifespan
```
We use the library(EpiDynamics) and the function SEIR() to implement a SEIR model:

<img src="http://www.public.asu.edu/~hnesse/classes/seireqn.png"/>  
<img src="https://upload.wikimedia.org/wikipedia/commons/3/3d/SEIR.PNG"/>

where the parameter beta here is time dependent, as estimated before.
```{r  scenario plot, echo=TRUE}
library(EpiDynamics)

# average number of single connections of an infected person
# less contacts, less probability of new infections
# we keep constant the other parameters
forecast<-14
seir1<-seir2<-seir3<-NULL
for(i in 1:forecast){
parameters <- c(mu = mu0, beta = beta.predict$fit[days+i], sigma = sigma0, gamma = 1/duration)
f1<-10
if( i==1) initials <- c(S = 0.95, E = (f1*I0/N), I = I0/N, R = R0/N)
if( i>1) initials <- c(S = seir1_temp$results$S[2], E = seir1_temp$results$E[2], I =seir1_temp$results$I[2], R = seir1_temp$results$R[2])
seir1_temp <- SEIR(pars = parameters, init = initials, time = 0:1)
seir1 <- rbind(seir1,SEIR(pars = parameters, init = initials, time = 0:1)$results[2,])
f2<-5
if( i==1) initials <- c(S = 0.95, E = (f2*I0/N), I = I0/N, R = R0/N)
if( i>1) initials <- c(S = seir2_temp$results$S[2], E = seir2_temp$results$E[2], I =seir2_temp$results$I[2], R = seir2_temp$results$R[2])
seir2_temp <- SEIR(pars = parameters, init = initials, time = 0:1)
seir2 <- rbind(seir2,SEIR(pars = parameters, init = initials, time = 0:1)$results[2,])
f3<-3
if( i==1) initials <- c(S = 0.95, E = (f3*I0/N), I = I0/N, R = R0/N)
if( i>1) initials <- c(S = seir3_temp$results$S[2], E = seir3_temp$results$E[2], I =seir3_temp$results$I[2], R = seir3_temp$results$R[2])
seir3_temp <- SEIR(pars = parameters, init = initials, time = 0:1)
seir3 <- rbind(seir3,SEIR(pars = parameters, init = initials, time = 0:1)$results[2,])
}

date<-seq(as.Date("2020-02-24"),as.Date("2020-02-24")+forecast-1+dim(dat_csv)[1],1)
plot(date,c(dat_csv$totale_attualmente_positivi,seir1$I*N),type="l",ylab="Cases",xlab="time",main="Infected")
lines(date,c(dat_csv$totale_attualmente_positivi,seir2$I*N),col=2)
lines(date,c(dat_csv$totale_attualmente_positivi,seir3$I*N),col=3)
lines(date[1:dim(dat_csv)[1]],dat_csv$totale_attualmente_positivi,lwd=2)
legend("topleft",c("first scenario","second scenario","third scenario"),lty=1,col=1:3)
```

The 3 scenarios show different numbers. 
If we consider the second scenario, at the end of the 2 weeks (`r date[length(date)]`) the number of infected is  (`r (seir2$I*N)[length(seir2$I)]`).

In the next plot the cumulative number of infected.  
At the end of the 2 weeks (`r date[length(date)]`) the total number of COVID19 cases is expected to be  (`r ((seir2$I+seir3$R)*N)[length(seir2$I)]`).

@@

plot(date,c(dat_csv$totale_casi,(seir1$I+seir1$R)*N),type="l",ylab="Cases",xlab="time",main="Cumulative Infected")
lines(date,c(dat_csv$totale_casi,(seir2$I+seir2$R)*N),col=2)
lines(date,c(dat_csv$totale_casi,(seir3$I+seir3$R)*N),col=3)
lines(date[1:dim(dat_csv)[1]],(dat_csv$totale_casi),lwd=2)
legend("topleft",c("first scenario","second scenario","third scenario"),lty=1,col=1:3)
```

### 2020-03-14 draft_analysis_Veneto.Rmd
Add home page, navigation and theme

---
title: "COVID19 - Forecast analysis"
author: "PG"
date: "3/11/2020"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
rm(list=ls())
###import italian dataset updated 10 March 2020
dat_csv<-read.csv("https://raw.githubusercontent.com/pcm-dpc/COVID-19/master/dati-regioni/dpc-covid19-ita-regioni.csv",header=T)
dat_csv<-dat_csv[dat_csv$codice_regione==5,]
dat_csv$t<-1:dim(dat_csv)[1]
plot(dat_csv$data,dat_csv$totale_attualmente_positivi,ylab="Infected",xlab="Date")
```

fit1 <- lm(log(totale_attualmente_positivi)~t,data=dat_csv)

```{r cars}
head(dat_csv)
fit1 <- lm(log(totale_attualmente_positivi)~t,data=dat_csv)
summary(fit1)
```

Estimates

```{r model plot, echo=TRUE}
plot(dat_csv$t,log(dat_csv$totale_attualmente_positivi),ylab="log cases",xlab="time")
abline(coef(summary(fit1))[,1])
```
The slope coefficient estimated in the linear regression model can be used to estimate R0, the number of 
```{r r0 plot, echo=TRUE}
slope <-coef(summary(fit1))[2,1]; slope
slope.se <- coef(summary(fit1))[2,2]; slope.se
### R0 estimates and 95%IC 
### I have used 14 for infection time, but it has a bimodal distribution (tested vs non tested)
R_0=slope*14+1;R_0
(slope+c(-1,1)*1.96*slope.se)*14+1
```
We want to make a short term forecast (14 days) with 3 scenario:  
-Scenario 1: 10 exposed people for each COVID-19 case and beta the same (no restrictions made or even no effects)
-Scenario 2: 10 exposed people for each COVID-19 case and beta reduced of 50% (-50% COVID19 contagious power)
-Scenario 3: 5 exposed people for each COVID-19 case and beta reduced of 50%  (-50% both exposed people and COVID19 contagious power)

We fix a series of initial parameters:
-I0: initial number of COVID-19 cases  
- R0: initial number of recovered  
- beta: the quantity connected to R0  
- N: Italian population   
- duration: infection duration of COVID-19   
- sigma0: the coronavirus transmission rate (half of flu epidemic)  
- mu0: the overall mortality rate  
```{r initation plot, echo=TRUE}
# initial number of infectus
I0<-dat_csv$totale_attualmente_positivi[dim(dat_csv)[1]]; I0
# initial number of recovered
R0<-dat_csv$dimessi_guariti[dim(dat_csv)[1]]; R0
#beta 
beta0<-R_0/(14)
# Veneto population
N=4905000
# duration of COVID19 
duration<-14
#sigma0 is the coronavirus transmission rate fixed to 5%  (half of flu epidemic)
sigma0<-0.05
#mortality rate 
mu0<-1/(82*365.25) # 1/lifespan
```
We use the library(EpiDynamics)
```{r first scenario plot, echo=TRUE}
library(EpiDynamics)
forecast<-14
parameters <- c(mu = mu0, beta = beta0, sigma = sigma0, gamma = 1/duration)
f1<-10
initials <- c(S = 0.95, E = (f1*I0/N), I = I0/N, R = R0/N)
seir1 <- SEIR(pars = parameters, init = initials, time = 0:forecast)
parameters <- c(mu = mu0, beta = beta0*1/2, sigma = sigma0, gamma = 1/duration)
f2<-10
initials <- c(S = 0.95, E = (f2*I0/N), I = I0/N, R = R0/N)
seir2 <- SEIR(pars = parameters, init = initials, time = 0:forecast)
parameters <- c(mu = mu0, beta = beta0*1/2, sigma = sigma0, gamma = 1/duration)
f3<-5
initials <- c(S = 0.95, E = (f3*I0/N), I = I0/N, R = R0/N)
seir3 <- SEIR(pars = parameters, init = initials, time = 0:forecast)

date<-seq(as.Date("2020-02-24"),as.Date("2020-02-24")+forecast-1+dim(dat_csv)[1],1)
plot(date,c(dat_csv$totale_attualmente_positivi,seir1$results$I[-1]*N),type="l",ylab="Cases",xlab="time",main="Infected")
lines(date,c(dat_csv$totale_attualmente_positivi,seir2$results$I[-1]*N),col=2)
lines(date,c(dat_csv$totale_attualmente_positivi,seir3$results$I[-1]*N),col=3)
lines(date[1:dim(dat_csv)[1]],dat_csv$totale_attualmente_positivi,lwd=2)
legend("topleft",c("first scenario","second scenario","third scenario"),lty=1,col=1:3)
```
The 3 scenarios show how measures of restriction can help to reduce the number of infected.  
At the end of the 2 weeks (`r date[length(date)]`) the number of infected is  (`r (seir2$results$I*N)[length(seir2$results$I)]`).
In the next plot the cumulative number of infected.
```{r  cumulative plot, echo=TRUE}

plot(date,cumsum(c(dat_csv$totale_attualmente_positivi,seir1$results$I[-1]*N)),type="l",ylab="Cases",xlab="time",main="Cumulative Infected")
lines(date,cumsum(c(dat_csv$totale_attualmente_positivi,seir2$results$I[-1]*N)),col=2)
lines(date,cumsum(c(dat_csv$totale_attualmente_positivi,seir3$results$I[-1]*N)),col=3)
lines(date[1:dim(dat_csv)[1]],cumsum(dat_csv$totale_attualmente_positivi),lwd=2)
legend("topleft",c("first scenario","second scenario","third scenario"),lty=1,col=1:3)
```

>>>>

---
title: "COVID19 - Forecast analysis - time dependent SEIR model - Veneto"
author: "PG"
date: "3/13/2020"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
rm(list=ls())
###import italian dataset updated 13 March 2020 - 
dat_csv<-read.csv("https://raw.githubusercontent.com/pcm-dpc/COVID-19/master/dati-regioni/dpc-covid19-ita-regioni.csv",header=T)
#we restricted the analysis to the Veneto Region
dat_csv<-dat_csv[dat_csv$codice_regione==5,]
dat_csv$t<-1:dim(dat_csv)[1]
plot(dat_csv$data,dat_csv$totale_attualmente_positivi,ylab="Infected",xlab="Date")
days<-dim(dat_csv)[1]
```

## The COVID dataset

The present analysis used the dataset on COVID19 updated in https://github.com/pcm-dpc/COVID-19.
We used a SEIR model to predict the trend of the actual COVID19 epidemic in Italy.
We estimated the R0 parameter by means of a linear regression model as reported in https://kingaa.github.io/clim-dis/parest/parest.html

The actual status is

```{r plots}
library(tidyr)
library(dplyr)
library(ggplot2)
df <- dat_csv %>%
  select(data, ricoverati_con_sintomi, terapia_intensiva, 
         totale_ospedalizzati, isolamento_domiciliare, 
         totale_attualmente_positivi, nuovi_attualmente_positivi, 
         dimessi_guariti, deceduti, totale_casi) %>%
  gather(key = "variable", value = "value", -data)
head(df, 3)

ggplot(df, aes(x = as.Date(data), y = value)) + 
  geom_line(aes(color = variable), size = 1) 
  ```

The plot shows an exponential grow. We estimate the R0 parameter by means of a linear model. R0 indicates how contagious an infectious disease is. It is also referred to as "the reproduction number" of COVID19.

#estimate r0
#see https://kingaa.github.io/clim-dis/parest/parest.html
#calculate r0 based in the last XX observation 
```{r R0 trend}
beta_vec<-NULL
sd_vec<-NULL
for (i in 1:(days-4)){
fit <- lm(log(totale_attualmente_positivi)~t,data=dat_csv[i:days,])
beta_vec<-c(beta_vec,coef(fit)[2])
sd_vec<-c(sd_vec,coef(summary(fit))[2,2])
}

label <-days:5
mean  <- (beta_vec*14)
lower <- ((beta_vec-1.96*sd_vec)*14)
upper <- ((beta_vec+1.96*sd_vec)*14)

df <- data.frame(label, mean, lower, upper)
df$label <- factor(df$label, levels=rev(df$label))

library(ggplot2)
fp <- ggplot(data=df, aes(x=label, y=mean, ymin=lower, ymax=upper)) +
  geom_pointrange() + 
  geom_hline(yintercept=1, lty=2) +  # add a dotted line at x=1 after flip
  coord_flip() +  # flip coordinates (puts labels on y axis)
  xlab("Last days used for the calculation") + ylab("R0 Mean (95% CI)") +
  theme_bw()  # use a white background
print(fp)

```

$$Y_t= a + \beta  t +e_t  $$

where $`Y_t`$ is the cumulative number of infected at the time t, while b is beta, the slope of the regression line.  
The slope coefficient estimated in the linear regression model can be used to estimate R0.

R0=slope*incubation period.

The incubation period for the coronavirus is in mean 5.1 days with a range from 2-14 days.
Please see https://www.worldometers.info/coronavirus/coronavirus-incubation-period/, but the incubation period is used for epidemic diseases that causes the immediate isolation of the infected subject.

In the calculation we considered an "incubation period" of 14 days for two reasons: 1) the majority of cases is asymptomatic, contagiousness is greater than 5, maybe 14. A minority (who made the swab) will have a duration of about 5 days between the start of contagiousness and swab; 2) 14 days is the worst scenario to consider in this phase.

We calculate several R0 values, each one based on a different number of days before the last day. 
The R0 shows a decreasing trend in the last period. The slope b indicates the rate of exponetial increase.  
However,R0 is going to decrease in the next days. We use the estimated trend between R0 and time to calculate the future R0 value for the next 14 days. 
We predict beta (and R0) for the next 14 days assuming a Gamma distribution for the beta (the slope) forcing its value greater than 0. 
```{r R0 forecast}
library(splines)
time<-1:(days-4)

beta.model<-glm(beta_vec~bs(time,3),weights = 1/sd_vec,family=Gamma)
forecast=14
# add 'fit', 'lwr', and 'upr' columns to dataframe (generated by predict)
pre<-predict(beta.model,type='response',newdata=data.frame(time=1:(days+forecast)),se.fit=TRUE)
date<-seq(as.Date("2020-02-24"),as.Date("2020-02-24")+forecast-1+dim(dat_csv)[1],1)
beta.predict <- data.frame(beta_vec=c(beta_vec,rep(NA,forecast+4)),time=date,fit=pre$fit,lwr=pre$fit-1*1.96*pre$se.fit,upr=pre$fit+1*1.96*pre$se.fit)

r0.predict<-beta.predict
r0.predict[,c(1,3:5)]<-r0.predict[,c(1,3:5)]*14
# plot the points (actual observations), regression line, and confidence interval
p <- ggplot(r0.predict, aes(date,beta_vec))
p <- p + geom_point() +labs(x="time from t0",y="R0 value") 
p <- p + geom_line(aes(date,fit))
p <- p + geom_ribbon(aes(ymin=lwr,ymax=upr), alpha=0.3)
p
```

R0 passes from values around to 3 in the initial phase to values <1 at the ending of the 14-days forecast.

We want to make a short term forecast (14 days) with 3 scenario:

-Scenario 1: 10 exposed people for each COVID-19 case (no restrictions made or even no effects)

-Scenario 2: 5 exposed people for each COVID-19 case (-50% exposed people)

-Scenario 3: 3 exposed people for each COVID-19 case  (-70% exposed people)

We use a SEIR model fixing a series of initial parameters:

- I0: initial number of COVID-19 cases  
- R0: initial number of recovered  
- beta: the quantity connected to R0 is considered to vary according the previous estimation
- N:  population   of the Veneto Region
- duration: infection duration of COVID-19   
- sigma0: the coronavirus transmission rate (half of flu epidemic)  
- mu0: the overall mortality rate

```{r initation plot, echo=TRUE}
# initial number of infectus
I0<-dat_csv$totale_attualmente_positivi[dim(dat_csv)[1]]; I0
# initial number of recovered
R0<-dat_csv$dimessi_guariti[dim(dat_csv)[1]]; R0
# italian poulation
N=4905854
# duration of COVID19 
duration<-14
#sigma0 is the coronavirus transmission rate fixed to 5%  (half of flu epidemic)
sigma0<-0.05
#mortality rate 
mu0<-1/(82*365.25) # 1/lifespan
```
We use the library(EpiDynamics) and the function SEIR() to implement a SEIR model:

<img src="http://www.public.asu.edu/~hnesse/classes/seireqn.png"/>  
<img src="https://upload.wikimedia.org/wikipedia/commons/3/3d/SEIR.PNG"/>

where the parameter beta here is time dependent, as estimated before.
```{r  scenario plot, echo=TRUE}
library(EpiDynamics)

# average number of single connections of an infected person
# less contacts, less probability of new infections
# we keep constant the other parameters
forecast<-14
seir1<-seir2<-seir3<-NULL
for(i in 1:forecast){
parameters <- c(mu = mu0, beta = beta.predict$fit[days+i], sigma = sigma0, gamma = 1/duration)
f1<-10
if( i==1) initials <- c(S = 0.95, E = (f1*I0/N), I = I0/N, R = R0/N)
if( i>1) initials <- c(S = seir1_temp$results$S[2], E = seir1_temp$results$E[2], I =seir1_temp$results$I[2], R = seir1_temp$results$R[2])
seir1_temp <- SEIR(pars = parameters, init = initials, time = 0:1)
seir1 <- rbind(seir1,SEIR(pars = parameters, init = initials, time = 0:1)$results[2,])
f2<-5
if( i==1) initials <- c(S = 0.95, E = (f2*I0/N), I = I0/N, R = R0/N)
if( i>1) initials <- c(S = seir2_temp$results$S[2], E = seir2_temp$results$E[2], I =seir2_temp$results$I[2], R = seir2_temp$results$R[2])
seir2_temp <- SEIR(pars = parameters, init = initials, time = 0:1)
seir2 <- rbind(seir2,SEIR(pars = parameters, init = initials, time = 0:1)$results[2,])
f3<-3
if( i==1) initials <- c(S = 0.95, E = (f3*I0/N), I = I0/N, R = R0/N)
if( i>1) initials <- c(S = seir3_temp$results$S[2], E = seir3_temp$results$E[2], I =seir3_temp$results$I[2], R = seir3_temp$results$R[2])
seir3_temp <- SEIR(pars = parameters, init = initials, time = 0:1)
seir3 <- rbind(seir3,SEIR(pars = parameters, init = initials, time = 0:1)$results[2,])
}

date<-seq(as.Date("2020-02-24"),as.Date("2020-02-24")+forecast-1+dim(dat_csv)[1],1)
plot(date,c(dat_csv$totale_attualmente_positivi,seir1$I*N),type="l",ylab="Cases",xlab="time",main="Infected")
lines(date,c(dat_csv$totale_attualmente_positivi,seir2$I*N),col=2)
lines(date,c(dat_csv$totale_attualmente_positivi,seir3$I*N),col=3)
lines(date[1:dim(dat_csv)[1]],dat_csv$totale_attualmente_positivi,lwd=2)
legend("topleft",c("first scenario","second scenario","third scenario"),lty=1,col=1:3)
```

The 3 scenarios show different numbers. 
If we consider the second scenario, at the end of the 2 weeks (`r date[length(date)]`) the number of infected is  (`r (seir2$I*N)[length(seir2$I)]`).

In the next plot the cumulative number of infected.  
At the end of the 2 weeks (`r date[length(date)]`) the total number of COVID19 cases is expected to be  (`r ((seir2$I+seir3$R)*N)[length(seir2$I)]`).

```{r  cumulative plot, echo=TRUE}

plot(date,c(dat_csv$totale_casi,(seir1$I+seir1$R)*N),type="l",ylab="Cases",xlab="time",main="Cumulative Infected")
lines(date,c(dat_csv$totale_casi,(seir2$I+seir2$R)*N),col=2)
lines(date,c(dat_csv$totale_casi,(seir3$I+seir3$R)*N),col=3)
lines(date[1:dim(dat_csv)[1]],(dat_csv$totale_casi),lwd=2)
legend("topleft",c("first scenario","second scenario","third scenario"),lty=1,col=1:3)
```
